---
title: "Data Challenge 2"
author: "Xueqing Zhou"
date: "2022-10-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE}
## load in the required libraries
library(readxl)
library(janitor)
library(stringr)
library(tidyverse)
library(lubridate)
library(hrbrthemes)
```


### **Problem 1**

```{r p1_ans, warning=FALSE, message=FALSE}
## read in the data
mr_trash_wheel <- read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
                          sheet = "Mr. Trash Wheel", range = cell_cols("A:N"))
prof_trash_wheel <- read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
                          sheet = "Professor Trash Wheel")
cap_trash_wheel <- read_excel("data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
                          sheet = "Captain Trash Wheel")

## wrangle the dataset `mr_trash_wheel`
mr_trash_wheel_clean <- mr_trash_wheel %>%
  clean_names() %>%  ## clean the column names
  filter(!is.na(dumpster)) %>%  ## omit rows that do not include dumpster-specific data
  mutate(wheel = "Mr. Trash Wheel") %>%  ## create new variable identifying data origin
  select(-c(date, volume_cubic_yards, homes_powered))  ## drop columns that are not needed

## wrangle the dataset `prof_trash_wheel`
prof_trash_wheel_clean <- prof_trash_wheel %>%
  clean_names() %>%  ## clean the column names
  filter(!is.na(dumpster)) %>%  ## omit rows that do not include dumpster-specific data
  mutate(wheel = "Professor Trash Wheel") %>%  ## create new variable identifying data origin
  select(-c(date, volume_cubic_yards, homes_powered))  ## drop columns that are not needed

## wrangle the dataset `cap_trash_wheel`
cap_trash_wheel_clean <- cap_trash_wheel %>%
  clean_names() %>%  ## clean the column names
  filter(!is.na(dumpster)) %>%  ## omit rows that do not include dumpster-specific data
  mutate(wheel = "Captain Trash Wheel") %>%  ## create new variable identifying data origin
  select(-c(date, volume_cubic_yards, homes_powered))  ## drop columns that are not needed

## create and wrangle the dataset `all_trash_wheels`
all_trash_wheels <- bind_rows(mr_trash_wheel_clean, 
                              prof_trash_wheel_clean, 
                              cap_trash_wheel_clean) %>%  ## bind three dataframes together
  pivot_longer(cols = c("plastic_bottles":"sports_balls", "plastic_bags"),
               names_to = "trash_type",
               values_to = "number") %>%  ## change to long format
  mutate(trash_type = str_to_title(str_replace(trash_type, "_", " ")))  ## format `trash_type`
  
## create and wrangle the dataset `all_trash_wheels_totals_June_2018`
all_trash_wheels_totals_June_2018 <- all_trash_wheels %>%
  filter(month == "June" & year == 2018) %>%
  group_by(wheel, trash_type) %>%
  summarise(total_num = sum(number))

## make a facted bar plot
all_trash_wheels_totals_June_2018 %>%
  ggplot(aes(total_num, wheel, fill = wheel)) +
  geom_bar(stat = "identity") +  ## make bar plot
  facet_wrap(~ trash_type, ncol = 2, scales = "free_x") +  ## facet by `trash_type`
  labs(x = "Amount of Trash",  ## relabel x axis
       y = "Trash Wheel Type",  ## relabel y axis
       title = "Amount of Trash Collected by Each Trash Wheel") +  ## add a title
  theme(legend.position = "none")  ## remove the legend
```


### **Problem 2**

```{r p2_ans, message=FALSE, warning=FALSE}
## read in the data
snp <- read_csv("data/snp.csv")
unemployment <- read_csv("data/unemployment.csv")

## wrangle the dataset `snp`
snp_clean <- snp %>%
  mutate(date = mdy(date),  ## convert to a date object
         date = if_else(year(date) > 2022, date - years(100), date),  ## correct the year problem
         year = year(date),  ## create a year variable
         month = month(date))  ## create a month variable

## wrangle the dataset `unemployment`
unemployment_clean <- unemployment %>%
  pivot_longer(!Year,
               names_to = "month",
               values_to = "rate") %>%  ## change to long format
  mutate(date = ymd(paste(Year, month, "01")))  ## create a `date` variable

## plot together
## value used to transform the data
coeff <- 150 

## define the color constant
snpColor <- "#69b3a2"
unemployColor <- rgb(0.2, 0.6, 0.9, 1)

ggplot() +
  geom_line(data = snp_clean, aes(date, close / coeff), color = snpColor) +  ## divide by `coeff` to get the same range
  geom_line(data = unemployment_clean, aes(date, rate), color = unemployColor) +  ## make line plot
  scale_y_continuous(name = "S&P Closing Price",  ## features of first axis
                     sec.axis = sec_axis(trans = ~.*coeff, name = "Unemployment Rate")) +  ## features of second axis
  theme_ipsum() +  ## remove the black background
  theme(axis.title.y = element_text(color = snpColor, size = 13),  ## customize the first y axis
        axis.title.y.right = element_text(color = unemployColor, size = 13)) +  ## customize the second y axis
  ggtitle("S&P Closing Price and Unemployment Rate Over Time")  ## add a title
```     
         

### **Problem 3**

```{r p3_ans, warning=FALSE}
## create a dataset `snp_average`
snp_average <- snp_clean %>%
  group_by(year, month) %>%  ## group by `year` and `month`
  mutate(mean_close = mean(close)) %>%  ## calculate the mean closing price
  mutate(date = ymd(paste(year, month, "01")))  ## create a `date` variable

## join two dataset and make a plot
merge(snp_average, unemployment_clean, by = "date") %>%  ## join two dataset together
  filter(date >= "2000-01-01") %>%  ## filter for data after the start of 2000
  ggplot(aes(mean_close, rate, color = as.factor(year))) +
  geom_point() +  ## make scatter plot
  labs(x = "S&P Mean Closing Price",  ## relabel x axis
       y = "Unemployment Rate",  ## relabel y axis
       title = "S&P Mean Closing Price Versus Unemployment Rate Over Time ",  ## add a title
       color = "Year")  ## relabel legend
```


### **Problem 4**

The plot in Problem 2 shows the S&P closing rate and unemployment rate over time from around 1950s to 2015. The S&P closing rate line increases over time, while the unemployment rate line fluctuates a lot and overall increases over time. For S&P closing rate, it has a steady increase from 1950 to 1990, and a rapid increase from 1990 to 2000. It fluctuates from 2000 to 2010, and results in increasing after 2010. For unemployment rate, it fluctuates over time. By comparing the start point in 1948 and the end point in 2015, there is a small overall rate increase.    

The plot in Problem 3 shows the S&P mean closing price versus unemployment rate over time. We can see two downward trends in the scatter plot. As S&P mean closing price increases, the unemployment rate decreases. From 2000 to 2004 and from 2005 to 2008, the S&P mean closing price decreases while the unemployment rate increases. From 2009 to 2015, the S&P mean closing price increases while the unemployment rate decreases. Overall looking, the unemployment rate reaches peak at a value of 10 in 
year 2009 with S&P mean closing price of 1050.

